Do music charts still matter in the streaming age? An analysis predicting Beyoncé’s Spotify numbers using a regression model.

### 2024-04-18 paper/paper.qmd
Add more references to paper
Fix a minor typo in paper
Fix minor typos and edit prose for final submission

The dataset used in this paper was created using data gathered from Spotify [@spotify] and Billboard [@billboard]. The data was collected by the authors for the purpose of this paper to investigate the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. Song titles, album titles, and the number of Spotify streams were collected from Spotify, while weeks spent on the *Billboard Hot 100* were collected from Billboard. There were no similar datasets available that could have been used.

@@

We analyzed the data in R [@citeR] using the following packages: `arrow` [@arrow], `broom.mixed` [@broommixed], `ggplot2` [@ggplot2], `here` [@here], `janitor` [@janitor], `knitr` [@knitr], `marginaleffects` [@marginaleffects], `modelsummary` [@modelsummary], `rstanarm` [@rstanarm], and `tidyverse` [@tidyverse].

@@

\begin{align} 
y_i|\mu_i,r &\sim \mbox{NegBinom}(\mu_i, r) \\
\mbox{log}(\mu_i) &= \alpha + \beta \times \mbox{Number of weeks}_i \\
\alpha &\sim \mbox{Normal}(18, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\end{align}

@@

Charts are supposed to reflect the industry’s reality and in 2018, it was announced that the *Billboard Hot 100* would take streaming data into account [@substack]. However, the exact formula behind Billboard’s calculations remains unknown to the public. We only know that one stream does not necessarily equal one sale and that some streams, such as for on demand services and from paid-tier users, count more than others [@substack]. By the 2010s, the question of what the Billboard charts actually reflect became even more prevalent after fans proved that it was susceptible to coordinated manipulation [@substack].

@@

Do artists as big as Beyoncé still care about music charts? Chart success can bring significant music promotion to an artist’s career. Songs that land on the *Billboard Hot 100* also land on Spotify’s biggest playlists, such as Today’s Top Hits which has almost 35 million followers [@todaystophits]. Cherie Hu, a music industry analyst at the intersection of music and data, has deemed playlists “the new album” and “more powerful than radio” as a placement on Today’s Top Hits can bring up to $163,000 in additional revenue per track [@cherie]. Moreover, for new artists, high positions in the charts can lead to more ownership and creative control in their musical career. This can help explain the importance of why many of Beyoncé’s longest charting songs come from her earlier albums. For more established artists who have a loyal fanbase, charting is essential for maintaining their brand and curtailing any image of a decline.

@@

Some weaknesses become apparent in our approach in using Spotify streams as a metric for the popularity of a Beyoncé song. First, one of Beyoncé’s albums, *Lemonade*, was added to Spotify three years after its release. This means that the number of Spotify streams for the songs from this album are underrepresented in our dataset. Furthermore, Spotify numbers are also not immune to industry politics that underlie promotional mechanics. All of the major record labels own a sizable share of Spotify and can give some artists preferential treatment on the platform. For example, Beyoncé is under Sony Music which owns a 2.65% stake in Spotify [@owners]. Lastly, we did not consider that Beyoncé’s older songs have had more time to accumulate streams than her latest work. For instance, *COWBOY CARTER* was just released on March 29, 2024 while *Dangerously in Love* was released on June 20, 2003 [@spotify].

In the future, it could be on interest to run the analysis again to give Beyoncé’s latest album, *COWBOY CARTER*, more time to see how long it lasts on the chart. We could also standardize the length to accumulate streams such as one year after its release for consistency. Another possible idea for future work could be to use a mix of new and established artists to see if we get the same result.

>>>>

The dataset used in this paper was created using data gathered from Spotify [@spotify] and Billboard [@billboard]. The data was collected by the author for the purpose of this paper to investigate the number of Spotify streams of a Beyoncé song based on the number of weeks it spent on the *Billboard Hot 100*. Song titles, album titles, and the number of Spotify streams were collected from Spotify, while weeks spent on the *Billboard Hot 100* were collected from Billboard. There were no similar datasets available that could have been used.

@@

We analyzed the data in R [@citeR] using the following packages: `arrow` [@arrow], `broom.mixed` [@broommixed], `ggplot2` [@ggplot2], `here` [@here], `janitor` [@janitor], `knitr` [@knitr], `marginaleffects` [@marginaleffects], `modelsummary` [@modelsummary], `rstanarm` [@rstanarm], and `tidyverse` [@tidyverse]. The code used in our analysis was adapted from *Telling Stories with Data* [@rohan].

@@

\begin{align} 
y_i|\mu_i,r &\sim \mbox{NegBinom}(\mu_i, r) \\
\mbox{log}(\mu_i) &= \alpha + \beta \times \mbox{Number of weeks}_i \\
\alpha &\sim \mbox{Normal}(18, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5)
\end{align}

@@

Charts are supposed to reflect the industry’s reality and in 2018, it was announced that the *Billboard Hot 100* would take streaming data into account [@substack]. However, the exact formula behind Billboard’s calculations remains unknown to the public. We only know that one stream does not necessarily equal one sale and that some streams, such as for on demand services and from paid-tier users, count more than others [@substack]. By the 2010s, the question of what the Billboard charts actually reflect became even more prevalent after fans proved that it was susceptible to coordinated manipulation [@substack]. Due to its ambiguity, we consider the effect of Billboard charts on a song’s popularity rather than what they represent.

@@

Do artists as big as Beyoncé still care about music charts? Chart success can bring significant music promotion to an artist’s career. Songs that land on the *Billboard Hot 100* also land on Spotify’s biggest playlists, such as *Today’s Top Hits* which has almost 35 million followers [@todaystophits]. Cherie Hu, a music industry analyst at the intersection of music and data, has deemed playlists “the new album” and “more powerful than radio” as a placement on *Today’s Top Hits* can bring up to $163,000 in additional revenue per track [@cherie].

Moreover, for new artists, high positions in the charts can lead to more ownership and creative control in their musical career. This can help explain the importance of why many of Beyoncé’s longest charting songs come from her earlier albums. For more established artists who have a loyal fanbase, charting is essential for maintaining their brand and curtailing any image of a decline.

@@

Some weaknesses become apparent in our approach in using Spotify streams as a metric for the popularity of a Beyoncé song. First, one of Beyoncé’s albums, *Lemonade*, was added to Spotify three years after its release. This means that the number of Spotify streams for the songs from this album are underrepresented in our dataset.

Furthermore, Spotify numbers are not immune to industry politics that underlie promotional mechanics. All of the major record labels own a sizable share of Spotify and can give some artists preferential treatment on the platform. For example, Beyoncé is under Sony Music which owns a 2.65% stake in Spotify [@owners].

Lastly, we did not consider that Spotify counts repeated streams and that Beyoncé’s older songs have had more time to accumulate streams. For instance, *COWBOY CARTER* was recently released on March 29, 2024 while *Dangerously in Love* was released on June 20, 2003 [@spotify].

In the future, it could be of interest to run the analysis again to give Beyoncé’s latest album, *COWBOY CARTER*, more time to see how long it lasts on the chart. We could also standardize the length of time to accumulate streams, such as one year after its release, for consistency. Another possible idea for future work could be to use a mix of new and established artists to see if we get the same result.
